Locally-Oriented Programming:\\ A Simple Programming Model for Stencil-Based Computations on Multi-Level Distributed Memory Architectures

### 2015-02-07 papers/LOPe/coarray-extensions.tex
Fixed one of Matt's \todo comments.
Paper lenght shortening begun.

We start with the declaration of an array (to be allocated later),
\begin{verbatim}
  real, allocatable, HALO(1:*:1,1:*:1),          &
        dimension(:,:), codimension[:,:] :: U
\end{verbatim}

with an explicit halo size and with two local dimensions (rank) and
two distributed memory codimensions (corank).  The corank of the array
is chosen to be identical to the rank of the array so that the logical
process topology aligns in a way that allows a natural halo exchange
between logically neighboring processes.\todo{What if corank and rank are not the same?}  For example, if the process
location is \texttt{[pcol,prow]}, then the right-hand halo for the
local array \texttt{U} can be obtained by the assignment
\texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size and cosize of
\texttt{U} are given by the allocation statement,

\begin{verbatim}
  allocate( U(0:M+1,0:N+1)[MP,*] )
\end{verbatim}

This allocation statement specifies (given the one element halo size
provided earlier for \texttt{U}) that the left halo column is
\texttt{U(0,:)}, the right column is \texttt{U(M+1,:)}, the bottom row
is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the
interior region \texttt{U(1:M,1:N)}.  In this allocation statement,
the total number of process rows \texttt{NP} can be obtained at
runtime, but \emph{may not} be explicitly provided (according to CAF
rules) because the actual number of participating processes (in
Fortran called images) is variable, depending on how many processes
are requested at program startup.  In this discussion, it is
\emph{assumed} that there are no holes in the logical processor
topology, thus $MP*NP = P$, where $P$ is the total number of
participating processes (images).

@@

\todo{subimages are critical and deserve their own subsection}

So far --- except for the \texttt{HALO} attribute --- this is totally
standard Fortran (2008).  For CAFe, we extend the concept of Fortran
images to be hierachical.  By this we mean that each image \emph{may}
have a subimage (or subimages), but this subimage is not visible to
other regular Fortran images.  Subimages also execute differently than
normal images.  Subimages are task based while images all execute a
single program but with different (multiple) data (SPMD).  A task can
be given to a subimage, but execution on the subimage terminates once
the task is finished.  Memory on a subimage is permanent, however, and
must be specifically allocated and deallocated.

One obtains a subimage by executing a function call,
\begin{verbatim}
   device = get_subimage(1)
\end{verbatim}

where the integer argument representing one of attached hardware
devices\todo{mention the relation of subimages and hardware earlier in
  defn. of subimage}.  If the function fails (e.g., the requested
device is unavailable) it returns the image number
\texttt{this\_image()} of the process that is executing the current
program.  Returning the current image allows program execution to
proceed correctly even if there are no attached devices.

@@

Execution of the \texttt{Laplace} task is done using the Fortran \texttt{do}
\texttt{concurrent} construct:
\begin{verbatim}
   do while (.not. converged)
      do concurrent (i=1:M, j=1:N)   [[device]]
         call Laplace( U(i,j)[device] )
      end do
      call HALO_TRANSFER(U, BC=CYCLIC)
   end do
\end{verbatim}
There are several points that require highlighting: 1. Iteration occurs over the interior
of the array domain, \texttt{(i=1:M, j=1:N)}; 2. Execution of the loop body occurs on the
specific subimage indicated by \texttt{[[device]]}; 3. Execution of the iterates may occur
in any order, even \emph{concurrently}; 4. The local element of the array (as
defined above in reference to the definition of the concurrent procedure
\texttt{Laplace}) is given by the indices \texttt{(i,j)}; 5. Location of memory for the
task is to be taken from the subimage as noted by \texttt{[device]}; 5. All threads must finish
execution of the loop body before further execution of the program proceeds; and 6. Transfer of
all requisite halo regions is effected by the call to the new LOPe intrinsic function
\texttt{HALO\_TRANSFER()}.  This function is a synchronization event in that all images must
complete the halo transfer before program execution continues.

@@

Finally, memory for the entire array \texttt{U} can be copied from the
subimage device with the statement,

\begin{verbatim}
  U = U[device]
\end{verbatim}

add\todo{Add?  is that typo?} memory deallocation is similar to memory allocation:

\begin{verbatim}
   deallocate(U)
   if (device /= this_image()) then
      deallocate(U[device])   [[device]]
   end if
\end{verbatim}

>>>>

We start with the declaration of an array (to be allocated later),
\begin{verbatim}
  real, allocatable, HALO(1:*:1,1:*:1),          &
        dimension(:,:), codimension[:,:] :: U
\end{verbatim}
with an explicit halo size and with two local dimensions (rank) and
two distributed memory codimensions (corank).  The corank of the array
is chosen to be identical to the rank of the array so that the logical
process topology aligns in a way that allows a natural halo exchange
between logically neighboring processes.\todo{What if corank and rank are not the same?}  For example, if the process
location is \texttt{[pcol,prow]}, then the right-hand halo for the
local array \texttt{U} can be obtained by the assignment
\texttt{U(M+1,:) = U(1,:)[pcol+1,prow]} where the size and cosize of
\texttt{U} are given by the allocation statement,
\texttt{allocate(U(0:M+1,0:N+1)[MP,*])}.
This allocation statement specifies (given the one element halo size
provided earlier for \texttt{U}) that the left halo column is
\texttt{U(0,:)}, the right column is \texttt{U(M+1,:)}, the bottom row
is \texttt{U(:,0)} and the top row is \texttt{U(:,N+1)}, leaving the
interior region \texttt{U(1:M,1:N)}.  In this allocation statement,
the total number of process rows \texttt{NP} can be obtained at
runtime, but \emph{may not} be explicitly provided (according to CAF
rules) because the actual number of participating processes (in
Fortran called images) is variable, depending on how many processes
are requested at program startup.  In this discussion, it is
\emph{assumed} that there are no holes in the logical processor
topology, thus $MP*NP = P$, where $P$ is the total number of
participating processes (images).

@@

So far --- except for the \texttt{HALO} attribute --- this is totally
standard Fortran (2008).  For CAFe, we extend the concept of Fortran
images to be hierachical.  By this we mean that each image \emph{may}
have a subimage (or subimages), but this subimage is not visible to
other regular Fortran images.  Subimages also execute differently than
normal images.  Subimages are task based while images all execute a
Single Program but with different (Multiple) Data (SPMD).  A task can
be given to a subimage, but execution on the subimage terminates once
the task is finished.  Memory on a subimage is permanent, however, and
must be specifically allocated and deallocated.

One obtains a subimage by executing a function call,
\texttt{device = GET\_SUBIMAGE(1)}
where the integer argument representing one of attached hardware
devices\todo{mention the relation of subimages and hardware earlier in
  defn. of subimage}.  If the function fails (e.g., the requested
device is unavailable) it returns the image number
\texttt{this\_image()} of the process that is executing the current
program.  Returning the current image allows program execution to
proceed correctly even if there are no attached devices.

@@

Execution of the \texttt{Laplace} task is done using the \texttt{do}
\texttt{concurrent} construct:
\begin{verbatim}
   do while (.not. converged)
      do concurrent (i=1:M, j=1:N)   [[device]]
         call Laplace( U(i,j)[device] )
      end do
      call HALO_TRANSFER(U, BC=CYCLIC)
   end do
\end{verbatim}
There are several points that require highlighting: 1. Iteration occurs over the interior
of the array domain, \texttt{(i=1:M, j=1:N)}; 2. Execution of the loop body occurs on the
specific subimage indicated by \texttt{[[device]]}; 3. Execution of the iterates may occur
in any order, even \emph{concurrently}; 4. The local element of the array (as
defined above in reference to the definition of the concurrent procedure
\texttt{Laplace}) is given by the indices \texttt{(i,j)}; 5. Location of memory for the
task is to be taken from the subimage as noted by \texttt{[device]}; 5. All threads must finish
execution of the loop body before further execution of the program proceeds; and 6. Transfer of
all requisite halo regions is effected by the call to the new LOPe intrinsic function
\texttt{HALO\_TRANSFER()}.  This function is a synchronization event in that all images must
complete the halo transfer before program execution continues.

@@

Finally, memory for the entire array \texttt{U} can be copied from the
subimage device with the statement,
\texttt{U = U[device]}, 
and memory deallocation (not shown) is similar to memory allocation.

%%\begin{verbatim}
%%   deallocate(U)
%%   if (device /= this_image()) then
%%      deallocate(U[device])   [[device]]
%%   end if
%%\end{verbatim}
